Supplementary Material for ``Engineering spin squeezing in a 3D optical lattice with interacting spin-orbit-coupled fermions''

### 2019-10-04 multi_body_interactions/papers/theory/paper.tex
minor cleanup

Second, \eqref{eq:H_int_fields} is only the first term in a low-energy
expansion of two-body interactions in effective field theory, which
generally includes additional terms containing derivatives of field
operators.  Derivative terms correspond to the dependence of two-body
scattering on the relative momentum $k$ of particles involved, with
$k\to0$ in the zero-energy limit.  In the present case of $s$-wave
scattering, the leading dependence of the two-body interaction
Hamiltonian on the relative momentum $k$ can be captured by use of an
energy-dependent pseudo-potential, which amounts to using a
$k$-dependent effective scattering length \cite{blume2002fermi}.  This
effective scattering length can be determined by expanding the
$s$-wave collisional phase shift in powers of the relative momentum
$k$ \cite{giorgini2008theory, flambaum1999analytical}.  Details of
this expansion will depend on the characteristic length scale of
finite-range interactions.  In our work, these corrections to
\eqref{eq:H_int_fields} will be relevant only for the calculation of
two-body interaction energies, appearing at third order in the
coupling constants $G_X$.  As we are primarily interested in $M$-body
interactions for $M\ge3$, we defer this calculation to Appendix
\ref{sec:momentum_dependence}.  We note that our approach of using an
unregularized contact potential, renormalizing coupling constants, and
separately accounting for momentum-dependent scattering is essentially
the same as the approach used for similar calculations in
refs.~\cite{johnson2012effective, yin2014universal,
  johnson2009effective}.  While this approach does not provide insight
into the microscopic structure of inter-atomic interactions, it is
suitable for the phenomenological description of these interactions,
and particular for our eventual development of a low-energy effective
theory.

>>>>

Second, \eqref{eq:H_int_fields} is only the first term in a low-energy
expansion of two-body interactions in effective field theory, which
generally includes additional terms containing derivatives of field
operators.  Derivative terms correspond to the dependence of two-body
scattering on the relative momentum $k$ of particles involved, with
$k\to0$ in the zero-energy limit.  In the present case of $s$-wave
scattering, the leading dependence of the two-body interaction
Hamiltonian on the relative momentum $k$ can be captured by use of an
energy-dependent pseudo-potential, which amounts to using a
$k$-dependent effective scattering length \cite{blume2002fermi}.  This
effective scattering length can be determined by expanding the
$s$-wave collisional phase shift in powers of the relative momentum
$k$ \cite{giorgini2008theory, flambaum1999analytical}.  Details of
this expansion will depend on the characteristic length scale of
finite-range interactions.  In our work, these corrections to
\eqref{eq:H_int_fields} will be relevant only for the calculation of
two-body interaction energies, appearing at third order in the
coupling constants $G_X$.  As we are primarily interested in $M$-body
interactions for $M\ge3$, we defer this calculation to Appendix
\ref{sec:momentum_dependence}.  We note that our approach of using an
unregularized contact potential, renormalizing coupling constants, and
separately accounting for momentum-dependent scattering is essentially
the same as the approach used for similar calculations in
refs.~\cite{johnson2012effective, yin2014universal,
  johnson2009effective}.  While this approach does not provide insight
into the microscopic structure of inter-atomic interactions, it is
suitable for the phenomenological description of these interactions,
and in particular for our eventual development of a low-energy
effective theory.

### 2019-10-04 soc_squeezing/collective_paper/collective_spin.tex
add graphics path (which was removed accidentally for submission)
minor cleanup

%%% figures
\usepackage{graphicx} % for figures
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\newcommand{\sref}[1]{\protect\subref{#1}} % for referencing subfigures

@@

Three primary considerations limit the maximum time $t$ to which we
can accurately compute a correlator $\bk{\hat S_{\v n}\p{t}}$ using
the TST expansion.  First, maintaining accuracy at larger times $t$
requires going to higher orders $M$ in the TST expansion.  An
order-$M$ TST expansion of the correlator $\bk{\hat S_{\v n}\p{t}}$
can involve a significant fraction of operators $\hat S_{\v m}$ with
weight $\abs{\v m}\lesssim M$, which implies the need to compute
$O\p{M^3}$ initial-time expectation values $\bk{\hat S_{\v m}}$ and
$O\p{M^4}$ matrix elements $T^k_{\v m\v n}$.  In practice, with a
straightforward implementation of the TST expansion we find that these
requirements generally restrict $M\lesssim 50$ -- $70$ with $8$ --
$50$ gigabytes of random access memory (RAM).  Second, individual
terms at high orders of the TST expansion in Eq.~\eqref{eq:TST} can
grow excessively large, greatly amplifying any numerical errors and
thereby spoiling cancellations that are necessary to arrive at a
physical value of a correlator, i.e.~with
$\abs{\bk{\hat S_{\v n}\p{t}}}\lesssim S^{\abs{\v n}}$.  Finally, the
TST expansion is essentially perturbative in the time $t$, which
implies that the validity of this a formal expansion eventually breaks
down.  Precisely characterizing the implications of these last two
considerations for the TST expansion requires additional analysis that
we defer to future work.  As we show from benchmarks of the TST
expansion in Section \ref{sec:squeezing}, however, a detailed
understanding of these limitations is not necessary to diagnose the
breakdown time $t_{\t{break}}^{(M)}$ beyond which the TST expansion
yields inaccurate results.  Empirically, we find that going beyond
order $M\approx35$ yields no significant gains in all cases considered
in this work.

@@

operator on $S$ with squared norm
$\norm{\O_S}^2\equiv\tr\p{\O_S^\dag\O_S}\le\tr\1_S$, then by linearity
we find that

>>>>

%%% figures
\usepackage{graphicx} % for figures
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\newcommand{\sref}[1]{\protect\subref{#1}} % for referencing subfigures
\graphicspath{{./figures/}} % set path for all figures

@@

Three primary considerations limit the maximum time $t$ to which we
can accurately compute a correlator $\bk{\hat S_{\v n}\p{t}}$ using
the TST expansion.  First, maintaining accuracy at larger times $t$
requires going to higher orders $M$ in the TST expansion.  An
order-$M$ TST expansion of the correlator $\bk{\hat S_{\v n}\p{t}}$
can involve a significant fraction of operators $\hat S_{\v m}$ with
weight $\abs{\v m}\lesssim M$, which implies the need to compute
$O\p{M^3}$ initial-time expectation values $\bk{\hat S_{\v m}}$ and
$O\p{M^4}$ matrix elements $T^k_{\v m\v n}$.  In practice, with a
straightforward implementation of the TST expansion we find that these
requirements generally restrict $M\lesssim 50$ -- $70$ with $8$ --
$50$ gigabytes of random access memory (RAM).  Second, individual
terms at high orders of the TST expansion in Eq.~\eqref{eq:TST} can
grow excessively large, greatly amplifying any numerical errors and
thereby spoiling cancellations that are necessary to arrive at a
physical value of a correlator, i.e.~with
$\abs{\bk{\hat S_{\v n}\p{t}}}\lesssim S^{\abs{\v n}}$.  Finally, the
TST expansion is essentially perturbative in the time $t$, which
implies that its validity as a formal expansion eventually breaks
down.  Precisely characterizing the implications of these last two
considerations for the TST expansion requires additional analysis that
we defer to future work.  As we show from benchmarks of the TST
expansion in Section \ref{sec:squeezing}, however, a detailed
understanding of these limitations is not necessary to diagnose the
breakdown time $t_{\t{break}}^{(M)}$ beyond which the TST expansion
yields inaccurate results.  Empirically, we find that going beyond
order $M\approx35$ yields no significant gains in all cases considered
in this work.

@@

operator on $S$ with operator norm $\norm{\O_S}\le1/\tr\1_S$
(i.e.~such that $\rho_S^{(2)}$ remains positive semi-definite, or a
valid quantum state), then by linearity we find that

### 2019-10-04 soc_squeezing/soc_squeezing_notes.tex
changed figure direcotry and set S \equiv N/2
minor cleanup

%%% figures
\usepackage{graphicx} % for figures
\usepackage{grffile} % help latex properly identify figure extensions
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\graphicspath{{./figures/}} % set path for all figures

@@

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_uniform} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}
  = \f{U}{L} \sum_{p,q}\f14\p{\1_q \1_p - \v\sigma_q\c\v\sigma_p},
  \label{eq:H_int_pauli}
\end{align}
with
\begin{align}
  \v{\sigma}_p \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z}.
\end{align}
For a fixed total particle number $N$, the identity operators in
\eqref{eq:H_int_pauli} contribute only a global shift in energy, which
allows us to more simply write
\begin{align}
  H_{\t{int}} = - \f{U}{L} \v S\c\v S,
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
\end{align}
This Hamiltonian has electronic eigenstates $\set{\ket{Sm}}$ with
total (pseudo-)spin $S$ and projection $m$ onto a quantization axis;
the corresponding energies are
$\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.  In particular, the
ground-state $S=N/2$ manifold is spanned by the Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m} \propto S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.

@@

manifold $\G_0$ of $H_0$, which have total spin $S=N/2$.  If the

@@

  H_{\t{eff}}

>>>>

%%% figures
\usepackage{graphicx} % for figures
\usepackage{grffile} % help latex properly identify figure extensions
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\graphicspath{{./figures/squeezing/}} % set path for all figures

@@

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}=\set{r,s}$; that is,
atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_uniform} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}
  = \f{U}{L} \sum_{p,q}\f14\p{\1_q \1_p - \v\sigma_q\c\v\sigma_p},
  \label{eq:H_int_pauli}
\end{align}
with
\begin{align}
  \v{\sigma}_p \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z}.
\end{align}
For a fixed total particle number $N$, the identity operators in
\eqref{eq:H_int_pauli} contribute only a global shift in energy, which
allows us to more simply write
\begin{align}
  H_{\t{int}} = - \f{U}{L} \v S\c\v S,
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
\end{align}
This Hamiltonian has electronic eigenstates $\set{\ket{Sm}}$ with
total (pseudo-)spin $S$ and projection $m$ onto a quantization axis;
the corresponding energies are
$\bk{Sm|H_{\t{int}}|Sm}=-\p{U/L}S\p{S+1}$.  In particular, the
ground-state $S=N/2$ manifold is spanned by the Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m} \propto S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.

@@

manifold $\G_0$ of $H_0$, which have total spin $S\equiv N/2$.  If the

@@

### 2019-10-04 soc_squeezing/squeezing_paper/supplementary.tex
minor cleanup

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{relative_error.pdf}
\caption{{\bf Relative error} between maximal squeezing (measured in dB) obtained by the OAT [Eqn.~(3)] and spin [Eqn.~(2)] models of the main text in a system of 20 particles.
The OAT model correctly captures the maximal squeezing (in dB) of the spin model to within 3\% (marked by the horizontal reference line) within the gap-protected regime $\widetilde{B}/U<0.06$.
}
\label{fig:relative_error}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{benchmarking_decay.pdf}
\caption{{\bf Comparison between the OAT and the spin model in the presence of decoherence.}
({\bf a}) The difference between the maximal squeezing (measured in dB) obtained by the OAT [Eqn.~(3)] and spin [Eqn.~(2)] models increases with the particle number $N$ and the single-particle spontaneous emission rate $\gamma$.
This disagreement is attributed in part to the fact that spontaneous emission transfers population of the collective spin state outside of the Dicke manifold, violating an assumption of the OAT model; see panel ({\bf b}).
The rate of population transfer outside of the Dicke manifold increases with both particle number and spontaneous emission rate.
(Parameters for simulations in this figure: $U=1000$ Hz, $J=200$ Hz, and $\phi=\pi/20$).
}
\label{fig:benchmarking_decay}
\end{figure*}

>>>>

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{relative_error.pdf}
\caption{{\bf Relative error} between maximal squeezing (measured in dB) obtained by the OAT [Eqn.~(4)] and spin [Eqn.~(3)] models of the main text in a system of 20 particles.
The OAT model correctly captures the maximal squeezing (in dB) of the spin model to within 3\% (marked by the horizontal reference line) within the gap-protected regime $\widetilde{B}/U<0.06$.
}
\label{fig:relative_error}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{benchmarking_decay.pdf}
\caption{{\bf Comparison between the OAT and the spin model in the presence of decoherence.}
({\bf a}) The difference between the maximal squeezing (measured in dB) obtained by the OAT [Eqn.~(4)] and spin [Eqn.~(3)] models increases with the particle number $N$ and the single-particle spontaneous emission rate $\gamma$.
This disagreement is attributed in part to the fact that spontaneous emission transfers population of the collective spin state outside of the Dicke manifold, violating an assumption of the OAT model; see panel ({\bf b}).
The rate of population transfer outside of the Dicke manifold increases with both particle number and spontaneous emission rate.
(Parameters for simulations in this figure: $U=1000$ Hz, $J=200$ Hz, and $\phi=\pi/20$).
}
\label{fig:benchmarking_decay}
\end{figure*}
